Isomorphisms as Groupoids: \\ Homotopy Type Theory with Fractional Types

### 2013-10-14 f.tex
0 types

\documentclass[11pt]{article}
\usepackage{fullpage}
\usepackage{url}
\usepackage{color, colortbl}
\usepackage{courier}
\usepackage{graphicx}

@@

A review of pi and slight reformulation of the semantics to use the groupoid
terminology. In more detail, motivate and introduce pointed types here, add
an element $\bot_A$ to every type $A$ and express the groupoids using a
pointed type and an explicit collection of discrete paths (not indexed by
anything at this point).

>>>>

\documentclass[11pt]{article}
\usepackage{fullpage}
\usepackage{url}
\usepackage{stmaryrd}
\usepackage{color, colortbl}
\usepackage{courier}
\usepackage{graphicx}

\newcommand{\alt}{~|~}
\newcommand{\leftv}[1]{\textsf{left}~#1}
\newcommand{\rightv}[1]{\textsf{right}~#1}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\identlp}{\mathit{identl}_+}
\newcommand{\identrp}{\mathit{identr}_+}
\newcommand{\swapp}{\mathit{swap}_+}
\newcommand{\assoclp}{\mathit{assocl}_+}
\newcommand{\assocrp}{\mathit{assocr}_+}
\newcommand{\identlt}{\mathit{identl}_*}
\newcommand{\identrt}{\mathit{identr}_*}
\newcommand{\swapt}{\mathit{swap}_*}
\newcommand{\assoclt}{\mathit{assocl}_*}
\newcommand{\assocrt}{\mathit{assocr}_*}
\newcommand{\distz}{\mathit{dist}_0}
\newcommand{\factorz}{\mathit{factor}_0}
\newcommand{\dist}{\mathit{dist}}
\newcommand{\factor}{\mathit{factor}}
\newcommand{\idc}{\mathit{id}}
\newcommand{\symc}[1]{\mathit{sym}~#1}
\newcommand{\proves}{\vdash}
\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2\\\end{array}}
{\begin{array}{l}#3\\\end{array}}$
 #4}}
\newcommand{\jdg}[3]{#2 \proves_{#1} #3}

@@

We review our language $\Pi$ providing the necessary background and context
for our higher-order extension.\footnote{The presentation in this section
  focuses on the simplest version of $\Pi$. Other versions include recursive
  types and trace operators but these extensions are orthogonal to the
  higher-order extension emphasized in this document.} Even though the
semantics of $\Pi$ could be explained using plain sets, we express it using
the groupoid terminology to faciliate the transition to fractionals and
higher-order functions.

%%%%%%%%%%%%%%%%%
\subsection{Syntax and Types}

The terms of $\Pi$ are not classical values and functions; rather, the terms
are isomorphism witnesses.  In other words, the terms of $\Pi$ are proofs
that certain ``shapes of values'' are isomorphic.  And, in classical
Curry-Howard fashion, our operational semantics shows how these proofs can be
directly interpreted as actions on ordinary values which effect this shape
transformation. Of course, ``shapes of values'' are very familiar already:
they are usually called \emph{types}.  But frequently one designs a type
system as a method of classifying terms, with the eventual purpose to show
that certain properties of well-typed terms hold, such as safety.  Our
approach is different: we start from a type system, and then present a term
language which naturally inhabits these types, along with an appropriate
operational and denotational semantics.

\paragraph*{Data.}
We view $\Pi$ as having two levels: it has traditional values classified by
ordinary types:
\[\begin{array}{lrcl} 
\textit{values}, v &::=& () \alt \leftv{v} \alt \rightv{v} \alt (v,v) \\
\textit{value types}, b &::=& 0 \alt 1 \alt b+b \alt b * b
\end{array}\]

Types include the empty type $0$, the unit type $1$, sum types $b_1+b_2$, and
product types $b_1*b_2$.  Values include $()$ which is the only value of type
$1$, $\leftv{v}$ and $\rightv{v}$ which inject $v$ into a sum type, and
$(v_1,v_2)$ which builds a value of product type.

\paragraph*{Isomorphisms.} The terms of $\Pi$ witness
type isomorphisms of the form $b \iso b$. They consist of base isomorphisms,
as defined below, and their composition.
\[\begin{array}{rrcll}
\identlp :&  0 + b & \iso & b &: \identrp \\
\swapp :&  b_1 + b_2 & \iso & b_2 + b_1 &: \swapp \\
\assoclp :&  b_1 + (b_2 + b_3) & \iso & (b_1 + b_2) + b_3 &: \assocrp \\
\identlt :&  1 * b & \iso & b &: \identrt \\
\swapt :&  b_1 * b_2 & \iso & b_2 * b_1 &: \swapt \\
\assoclt :&  b_1 * (b_2 * b_3) & \iso & (b_1 * b_2) * b_3 &: \assocrt \\
\distz :&~ 0 * b & \iso & 0 &: \factorz \\
\dist :&~ (b_1 + b_2) * b_3 & \iso & (b_1 * b_3) + (b_2 * b_3)~ &: \factor
\end{array}\]

Each line of the above table introduces a pair of dual
constants\footnote{where $\swapp$ and $swapt$ are self-dual.} that witness
the type isomorphism in the middle.  These are the base (non-reducible) terms
of the second, principal level of $\Pi$. Note how the above has two readings:
first as a set of typing relations for a set of constants. Second, if these
axioms are seen as universally quantified, orientable statements, they also
induce transformations of the (traditional) values. The (categorical or
homotopical) intuition here is that these axioms have computational content
because they witness isomorphisms rather than merely stating an extensional
equality.

The isomorphisms are extended to form a congruence relation by adding the
following constructors that witness equivalence and compatible closure:

\begin{table}[t]
\hrule
\smallskip
\begin{center} 
\Rule{}
{}
{\jdg{}{}{\idc : b \iso b}}
{}
\quad 
\Rule{}
{\jdg{}{}{c : b_1 \iso b_2}}
{\jdg{}{}{\symc{c} : b_2 \iso b_1}}
{}
\quad
\Rule{}
{\jdg{}{}{c_1 : b_1 \iso b_2} \quad c_2 : b_2 \iso b_3}
{\jdg{}{}{c_1 \fatsemi c_2 : b_1 \iso b_3}}
{}
\medskip 
\Rule{}
{\jdg{}{}{c_1 : b_1 \iso b_2} \quad c_2 : b_3 \iso b_4}
{\jdg{}{}{c_1 \oplus c_2 : b_1 + b_3 \iso b_2 + b_4}}
{}
\quad 
\Rule{}
{\jdg{}{}{c_1 : b_1 \iso b_2} \quad c_2 : b_3 \iso b_4}
{\jdg{}{}{c_1 \otimes c_2 : b_1 * b_3 \iso b_2 * b_4}}
{}
\end{center}
\caption{Combinators\label{pi-combinators}}
\smallskip
\hrule
\end{table}

It is important to note that ``values'' and ``isomorphisms'' are completely
separate syntactic categories which do not intermix. The semantics of the
language come when these are made to interact at the ``top level'' via
\emph{application}: 
\[\begin{array}{lrcl}
\textit{top level term}, l &::=& c v
\end{array}\]

The language presented above, at the type level, models a \emph{semiring}
(occasionally called a \emph{rig}) where we replace equality by isomorphism.
Semantically, $\Pi$ models a \emph{bimonoidal category} whose simplest
example is the category of finite sets and bijections. In that
interpretation, each value type denotes a finite set of a size calculated by
viewing the types as natural numbers and each combinator $c : b_1 \iso b_2$
denotes a bijection between the sets denoted by $b_1$ and $b_2$.

%%%%%%%%%%%%%%%%%
\subsection{Operational Semantics}

Operationally, the semantics of $\Pi$ is given using two mutually recursive
interpreters: one going forward and one going backwards. The use of $\symc{}$
switches control from one evaluator to the other.  We state without proof
that the evaluation of well-typed combinators always terminates and that
$\Pi$ is logically reversible, i.e., that for all combinators $c : b_1 \iso
b_2$ and values $v_1 : b_1$ and $v_2 : b_2$ we have the forward evaluation of
$c v_1$ produces $v_2$ iff the backwards evaluation of $c v_2$ produces
$v_1$.

%%%%%%%%%%%%%%%%%
\subsection{Groupoid Semantics}

In more detail, motivate and introduce pointed types here, add
an element $\bot_A$ to every type $A$ and express the groupoids using a
pointed type and an explicit collection of discrete paths (not indexed by
anything at this point).
